Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-07-26 manuscript/usses_water_ms.tex
rewriting discussion

\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, Utah 84322}
\affil[2]{Department of Ecology and Evolutionary Biology, University of
California, Los Angeles, Los Angeles, California 90095}
\affil[3]{USSES, address\ldots{}}

@@

Projected changes in precipitation regimes, associated with global
climate change, will affect

Water availability is a strong driver of aboveground net primary
productivity (ANPP) in grassland ecosystems. Mean annual precipitation
(MAP) is strongly correlated with ANPP across sites \citep{Huxman2004},
and, in water-limited ecosystems, interannual variation in MAP is also
correlated with interannual variation in ANPP \citep{Hsu2012}. These
relationships suggest that chronic alterations in levels of
precipitation, associated with global climate change, will impact ANPP.

At a given site, the functional response of ANPP to water availability
(e.g., soil moisture) can be characterized by fitting a model to
historical observations of ANPP and soil moisture. However, the fitted
functional response may provide an incomplete picture because future
conditions are likely to be outside the historical range of variability
\citep{Smith2011}. For example, historical trends may show a linear soil
moisture-ANPP relationship, but that relationship could potentially
saturate if soil moisture is pushed far beyond typical levels.
Saturating relationships are actually common
\citep{Hsu2012, Gherardi2015a}, perhaps because other resources, like
nitrogen, become more limiting in wet years than dry years. Knowing the
curvature of the soil moisture-ANPP relationship is critical for
understanding how ecosystems will respond to chronic alterations in
water availability.

@@

Manipulating potentially limiting resources like precipitation offers a
route to understanding how ecosystems will respond to resource levels
that fall outside the historical range of variability
\citep{Avolio2015, Gherardi2015, Knapp2017}. Altering the amount of
precipitation over many years should provide insight into the time
scales at which water-limited ecosystems respond to chronic resource
alteration. Following the HRF, we propose four alternative predictions
for the effect of precipitation manipulation on the ecosystem functional
response to soil moisture, that is, the soil moisture-ANPP relationship
(Figure 1). The four predictions are based on possible outcomes at the
community (e.g., community composition) and ecosystem (e.g., soil
moisture-ANPP regression) levels.

@@

Between 1926 and 1932, range scientists at the USSES established 26
permanent 1 m\(^2\) quadrats to track vegetation change over time. In
2007, we (well, one of us {[}P. Adler{]}) relocated 14 of the original
quadrats, six of which were inside a large, permanent livestock
exclosure. We use these six plots as control plots that have recieved no
treatment, just ambient precipitation. In spring 2011, we (well, two of
us {[}A. Kleinhesselink and P. Adler{]}) established 16 new 1 m\(^2\)
plots located in one of the same exclosures as 6 of the original
vegetation monitoring plots. We avoided areas on steep hill slopes,
areas with greater than 20\% cover of bare rock, and areas with greater
than 10\% cover of the shrubs \emph{Purshia tridentata} and/or
\emph{Amelanchier utahensis}. We established the new plots in pairs and
randomly assigned each plot in a pair to receive a ``drought'' or
``irrigation'' treatment.

@@

Analyzing the response to experimental treatments was complicated by the
fact that we did not directly monitor soil moisture in each plot on each
day of the experiment.\\
Only a subset of plots were equiped with soil moisture sensors, and
within those plots, one or more of the sensors frequently failed to
collect data. To remedy these problems, and to produce average daily
soil moisture values for the ambient, drought, and irrigation
conditions, we used a statistical model to predict the average treatment
effects on soil moisture during the course of the experiment.

@@

We then modeled the daily deviation from ambient conditions of the
drought and irrigation treatments using a linear mixed effects model
with independent variables for treatment, season (winter, spring,
summer, fall), rainfall, and all two-way interactions.\\
Rainy days were defined as any day in which precipitation was recorded
and average temperature was above 3\(^{\circ}\)C. The day immediately
following rainfall was also classified as rainy. We fit the model using
the `lme4::lmer()' function \citep{Bates2015} in R \citep{R2016}, with
random effects for plot group and date. We weighted observations by the
number of unique sensors or spot measurements that were taken in each
plot on that day. We then used the model to predict the average daily
soil moisture in the treated plots based on the average daily ambient
soil moisture.\\
We could only predict soil moisture in the treated plots on days for
which we took at least one ambient soil moisture measurement (Fig. 2B).

@@

Cumulative March-June soil moisture had a weak positive effect on ANPP
(mean of \(\beta_{1}\) = 0.41), but the effect is associated with high
uncertainty, with credible intervals that overlap zerp (80\% BCI =
-0.08, 0.86; 95\% BCI = -0.45, 1.13) (Fig. 3). Ecosystem functional
response was similar among treatments, with treatment level regressions
between soil moisture and log(ANPP) having similar intercepts and slopes
(Fig. 3A). There was also no evidence that the treatment effects became
more important over time because there was no directional trend in the
random year effects (Fig. 4).

@@

Ecosystem response to chronic resource alteration is expected to follow
a temporal trend. Initially, ecosystem response will be modest and
reflect the physiological responses of constituent species. Over longer
time periods, species reordering will cause greater responses as species
better able to take advantage of, or cope with, new resource levels
become more abundant. This temporal trend is formalized by the
`Hierarchical Response Framework' \citep[HRF,][]{Smith2009}, and has
been empirically supported \citep{Knapp2012, Wilcox2016}. A lingering
question is not whether the HRF is a reasonable model of ecosystem
dynamics, but rather how the time scales of ecosystem response and
community change differ among ecosystems. Indeed, previous work has
shown that community compositional shifts can be both rapid, on the
order of years \citep{Hoover2014}, and slow, on order of decades
\citep{Knapp2012, Wilcox2016}. To add to this growing body of knowledge,
we performed a precipitation manipulation experiment in a sagebrush
steppe ecosystem.

Ecosystem functional response (i.e., the relationship between soil
moisture and ANPP) and community composition in this sagebrush steppe
ecosystem was consistent across drought, control, and irrigation
treatments (Fig. 2D, Fig. 3). In other words, functional responses
estimated from any precipitation level sampled apply to any other level
of precipitation, and the consistent functional response was not due to
species reordering and compensation. Thus, our results conform to the
first of our four predictions: neither ecosystem functional response nor
community composition changed with chronic alteration in water
availability (Fig. 1, top left).

This is surprising because grasslands generally, and sagebrush steppe
specifically, are considered water-limited systems. Thus, our naive
expecatations were that either (i) ecosystem functional response would
differ among treatments as species reach physiological thresholds and
resource limitations (Fig. 1, top right), or (ii) ecosystem functional
response stays the same due to species reordering and compensation (Fig.
1, bottom left).

In the absence of community composition data, an obvious hypothesis to
explain the consistent ecosystem functional response is that species
reordered rapidly. In so doing, newly abundant species could compensate
for the loss of ecosystem functioning of previously abundant species. We
did not find this to be the case, as species composition was remarkably
stable over the course of the experiment (Fig. 5).

In combination, the lack of ecosystem and community response to drought
or irrigation shows that this sagebrush steppe ecosystem is resistant to
chronic alterations in water availability. There are three possible
explanations for the resistance we found. First, it could be that our
experiment simply was not long enough to induce responses. This may be
true for community responses, which can take decades \citep{Wilcox2016}.

Second, it could be that our manipulations were not large enough to
induce a response. That is, maybe a 50\% decrease/increase in any given
year is not abnormal give our sites historical range of variability
\citep{Knapp2017}. We cannot definitively rule out this possibility, but
we have reason to believe our manipulations \emph{should} have been
large enough. Using the methods described by \citet{Lemoine2016}, we
calculated the percent reduction and increase of mean growing season
precipitation neccesary to reach the 1\% and 99\% extremes of the
historical precipitation regime at our site. The 1\% quantile of
precipitation at our site is 110 mm, a 47\% reduction from the mean, and
the 99\% quantile is 414 mm, a 77\% increase from mean growing season
precipitation (Appendix 3). Thus, our drought treatment represented
extreme precipitation amounts, especially in years where ambient
precipitation was below average (Fig. 2A). The irrigation treatment may
have been too small, however.

@@

In conclusion, our results suggest that five years of \(\pm\) 50\%
ambient precipitation is not enough to induce a shift in ecosystem
functional response in a sagebrush steppe. This is likely because
component species are tolerant to drought conditions and water remains
co-limiting at higher precipitation levels. Longer time series of
chronic precipitation alteration may reveal plant community shifts that
we did not observe \citep[e.g.,][]{Wilcox2016}. Our results suggest
compositional shifts would have the largest impact at high rainfall
because the current community maintained consistent ecosystem functional
response at very low water availability.

@@

NSF DBI-1400370 to Andrew Tredennick.\\
NSF Graduate Research Fellowship to Andrew Kleinhesselink.\\
NSF DEB-1353078 and DEB-1054040 to Peter Adler.

@@

\begin{itemize}
  \item Andrew T. Tredennick collected data, analyzed the data, wrote the paper, prepared figures and/or tables, reviewed
drafts of the paper.
  \item Andrew R. Kleinhesselink conceived and designed the experiments, performed the experiments, collected data, reviewed drafts of the paper.
  \item Peter B. Adler conceived and designed the experiments, performed the experiments, collected data, analyzed the data, reviewed
drafts of the paper.
\end{itemize}

>>>>

\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, Utah 84322}
\affil[2]{Department of Ecology and Evolutionary Biology, University of
California, Los Angeles, Los Angeles, California 90095}
\affil[3]{United States Department of Agriculture, Agriculture Research Station,
U.S. Sheep Experiment Station, Dubois, Idaho 83423}

@@

Ecosystems worldwide are projected to experience either wetter or drier
conditions in the future \citep{Zhang2007, Greve2014}. This has
important implications for ecosystem functioning because mean annual net
primary productivity (ANPP) positively covaries with mean annual
precipitation across ecosystems \citep{Sala1988, Huxman2004} and
interannual responses of ANPP are linked to interannual variability in
precipitation within ecosystems
\citep{Lauenroth1992, Hsu2012, Gherardi2015a}. To predict how ecosystems
might respond to extreme precipitation levels, we need to understand
whether ecosystems tend to follow temporal trends, which assume
relatively constant plant community composition, or spatial trends,
which allow community composition to shift \citep{Wilcox2016}.

At a given site, the functional response of ANPP to water availability
(e.g., soil moisture) can be characterized by fitting a model to
historical observations of ANPP and soil moisture. However, the fitted
functional response may provide an incomplete picture because future
conditions are likely to be outside the historical range of variability
\citep{Smith2011}. For example, historical trends may underestimate the
potential for the soil moisture-ANPP relationship to saturate if soil
moisture is pushed far beyond typical levels. Saturating relationships
are actually common \citep{Hsu2012, Gherardi2015a}, perhaps because
other resources, like nitrogen, become more limiting in wet years than
dry years. Knowing the curvature of the soil moisture-ANPP relationship
at extreme precipitation levels is critical for understanding how
ecosystems will respond to chronic alterations in water availability.

@@

Manipulating potentially limiting resources, like precipitation, offers
a route to understanding how ecosystems will respond to resource levels
that fall outside the historical range of variability
\citep{Avolio2015, Gherardi2015, Knapp2017}. Altering the amount of
precipitation over many years should provide insight into the time
scales at which water-limited ecosystems respond to chronic resource
alteration. Following the HRF, we propose four alternative predictions
for the effect of precipitation manipulation on the ecosystem functional
response to soil moisture, that is, the soil moisture-ANPP relationship
(Fig. 1). The four predictions are based on possible outcomes at the
community (e.g., community composition) and ecosystem (e.g., soil
moisture-ANPP regression) levels.

@@

Between 1926 and 1932, range scientists at the USSES established 26
permanent 1 m\(^2\) quadrats to track vegetation change over time. In
2007, we (well, one of us {[}P. Adler{]}) relocated 14 of the original
quadrats, six of which were inside a large, permanent livestock
exclosure. We use these six plots as control plots that have recieved no
treatment, just ambient precipitation. In spring 2011, we (well, two of
us {[}A. Kleinhesselink and P. Adler{]}) established 16 new 1 m\(^2\)
plots located in the same exclosure as the six control plots. We avoided
areas on steep hill slopes, areas with greater than 20\% cover of bare
rock, and areas with greater than 10\% cover of the shrubs \emph{Purshia
tridentata} and/or \emph{Amelanchier utahensis}. We established the new
plots in pairs and randomly assigned each plot in a pair to receive a
``drought'' or ``irrigation'' treatment.

@@

Analyzing the response to experimental treatments was complicated by the
fact that we did not directly monitor soil moisture in each plot on each
day of the experiment. Only a subset of plots were equiped with soil
moisture sensors, and within those plots, one or more of the sensors
frequently failed to collect data. To remedy these problems, and to
produce average daily soil moisture values for the ambient, drought, and
irrigation conditions, we used a statistical model to predict the
average treatment effects on soil moisture during the course of the
experiment.

@@

We then modeled the daily deviation from ambient conditions of the
drought and irrigation treatments using a linear mixed effects model
with independent variables for treatment, season (winter, spring,
summer, fall), rainfall, and all two-way interactions. Rainy days were
defined as any day in which precipitation was recorded and average
temperature was above 3\(^{\circ}\)C. The day immediately following
rainfall was also classified as rainy. We fit the model using the
`lme4::lmer()' function \citep{Bates2015} in R \citep{R2016}, with
random effects for plot group and date. We weighted observations by the
number of unique sensors or spot measurements that were taken in each
plot on that day. We then used the model to predict the average daily
soil moisture in the treated plots based on the average daily ambient
soil moisture. We could only predict soil moisture in the treated plots
on days for which we took at least one ambient soil moisture measurement
(Fig. 2B).

@@

Cumulative March-June soil moisture had a weak positive effect on ANPP
(mean of \(\beta_{1}\) = 0.41; Fig. 3), but the effect is associated
with high uncertainty, with credible intervals that overlap zerp (80\%
BCI = -0.08, 0.86; 95\% BCI = -0.45, 1.13). Ecosystem functional
response was similar among treatments, with treatment level regressions
between soil moisture and log(ANPP) having similar intercepts and slopes
(Fig. 3A). There was also no evidence that the treatment effects became
more important over time because there was no directional trend in the
random year effects (Fig. 4).

@@

Ecosystem response to precipitation extremes depends on the
physiological responses of constituent species and the rate at which
community composition shifts to favor species better able to take
advantage of, or cope with, new resource levels \citep{Smith2009}.
Previous work has shown that community compositional shifts can be both
rapid, on the order of years \citep{Hoover2014}, and slow, on order of
decades \citep{Knapp2012, Wilcox2016}. Thus, a lingering question is how
the time scales of ecosystem response and community change differ among
ecosystems, which can be answered by manipulating precipitation to reach
extreme levels. The results of our five year experiment in a sagebrush
steppe conform to the first of our four predictions: neither ecosystem
functional response nor community composition changed with chronic
alteration in water availability (Fig. 1, top left).

The similarity of ecosystem functional response (Fig. 3A) and community
composition (Fig. 5) across treatments is surprising because grasslands
generally, and sagebrush steppe specifically, are considered
water-limited systems. Indeed, we expected ecosystem functional
response, community composition, or both to change with precipitation
treatment, landing us in any box of Fig. 1 \emph{except} the top left.
So, why did our treatments fail to induce ecosystem or community
responses? We can think of three reasons; two are limitations of our
study, and one is the life history traits of the species in our focal
communities. We first discuss the potential limitations of our study,
and then discuss biological explanations.

First, it could be that our manipulations were not large enough to
induce a response. That is, maybe a 50\% decrease/increase in any given
year is not abnormal given our site's historical range of variability
\citep{Knapp2017}. We cannot definitively rule out this possibility, but
we have reason to believe our manipulations \emph{should} have been
large enough. Using the methods described by \citet{Lemoine2016}, we
calculated the percent reduction and increase of mean growing season
precipitation neccesary to reach the 1\% and 99\% extremes of the
historical precipitation regime at our site. The 1\% quantile of
precipitation at our site is 110 mm, a 47\% reduction from the mean, and
the 99\% quantile is 414 mm, a 77\% increase from mean growing season
precipitation (Appendix 3). Thus, our drought treatment represented
extreme precipitation amounts, especially in years where ambient
precipitation was below average (Fig. 2A). The irrigation treatment may
have been too small, however.

Second,

@@

In conclusion, our results suggest that five years of \(\pm\) 50\%
ambient precipitation is not enough to induce a shift in ecosystem
functional response in a sagebrush steppe. This is likely because the
species in our focal plant community are bet hedgers, maintaining
relatively consistent responses to interannual variation in
precipitation to avoid booms and busts. Longer time series of chronic
precipitation alteration may reveal plant community shifts that we did
not observe \citep[e.g.,][]{Wilcox2016}, in which case species that do
not bet hedge may gain prominence and dominate the ecosystem functional
response. Our results suggest compositional shifts would have the
largest impact at high rainfall because the current community maintained
consistent ecosystem functional response at very low water availability.

@@

NSF DBI-1400370 to Andrew T. Tredennick.\\
NSF Graduate Research Fellowship to Andrew R. Kleinhesselink.\\
NSF DEB-1353078 and DEB-1054040 to Peter B. Adler.

@@

\begin{itemize}
  \item Andrew T. Tredennick collected data, analyzed the data, wrote the paper, prepared figures and/or tables, reviewed
drafts of the paper.
  \item Andrew R. Kleinhesselink conceived and designed the experiments, performed the experiments, collected data, reviewed drafts of the paper.
  \item Bret Taylor...
  \item Peter B. Adler conceived and designed the experiments, performed the experiments, collected data, analyzed the data, reviewed
drafts of the paper.
\end{itemize}
